---
title: "Portfolio"
description: |
  Check out some of my work!
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# attach packages
library(tidyverse)
library(here)
library(janitor)
library(plotly)
library(ggrepel)
```

```{r}
# read in data using here() to create a system agnostic file path

mono_lake <- read_csv(here("data", "mono_lake_annual_levels.csv"), skip = 5) %>% # use skip to begin the table at a different row than row 1
  clean_names() # make column names coder friendly
```

```{r}
mono_history <- mono_lake %>% # create a dataframe with years of historical events
  filter(year %in% c(1941, 1976, 1978, 1994, 2013)) %>% # filter for relevant years
  mutate(significance = c("diversion begins", "initial ecological study", "MLC founded", "stability level est.", "MBSRA approved")) # add column to attribute historical significance
```


```{r}
ggplot(data = mono_lake, aes(x = year, y = lake_level_feet_above_sea_level)) + # create a plot using the complete mono_lake data
  geom_line(color = "deepskyblue2", size = 1) + # make it a blue line graph
  labs(y = "feet above sea level", # label axes and title
       title = "Mono Lake Annual Levels") +
  geom_hline(yintercept = 6392, color = "darkolivegreen4", alpha = 0.7) + # add horizontal lines for significant lake levels
  geom_hline(yintercept = 6377, color = "goldenrod1", alpha = 0.7) +
  geom_hline(yintercept = 6360, color = "coral1", alpha = 0.7) +
  geom_point(data = mono_history, # create overlapping plot with mono_history data frame
             aes (x = year, y = lake_level_feet_above_sea_level), # assign x and y variables
             color = "darkorchid1", size = 1.25) + # add color and change size of points
  theme(plot.title = element_text(hjust = 0.5)) + # add a plot title and center it
  annotate("text", # add label for horizontal lines of significant levels
           y = 6390, # dictate location
           x = 1862,
           size = 3, # dictate size
           label = "stable lake level", # label the line
           color = "darkolivegreen4") + # add color
  annotate("text", # do same as above for each line
           y = 6375,
           x = 1865.5,
           size = 3,
           label = "land bridge emerges",
           color = "goldenrod1") +
  annotate("text", # do same as above for each line
           y = 6358,
           x = 1874.5,
           size = 3,
           label = "salinity impedes brine shrimp pop",
           color = "coral1") +
  geom_label_repel(data = mono_history, # add labels to the years of historical significance
                   aes(x = year, y = lake_level_feet_above_sea_level, # indicate x and y variables
                       label = significance), # label the points according to the significance column
                   box.padding = 1, # box. and point.padding adjust position and size of labels
                   point.padding = 0.1) +
  scale_x_continuous(limits = c(1850, 2017), # adjust x axis scale to be from 1850 to 2020
                     breaks = seq(1850, 2020, 10), # breaks every 10 years
                     expand = c(0, 0)) + # remove buffer zone
  scale_y_continuous(limits = c(6350, 6430), # adjust y axis scale to be from 6350 to 6430
                     breaks = seq(6350, 6430, 20), # breaks every 20 feet
                     expand = c(0, 0)) # remove buffer zone
```

*Figure 1*: Mono Lake Annual Levels. Annual levels of Mono Lake from 1850 to 2017 in terms of feet above sea level. Since the 1940s, Los Angeles's Department of Water & Power has been diverting water away from Mono Lake in order to meet the city's water demands. A study in the 1970s revealed that reducing the level and volume of the lake was resulting in ecological damage affecting the native wildlife including California Gulls, Eared Grebes, Osprey, and Brine Shrimp, among many others. The students who conducted the study founded the Mono Lake Committee (MLC) with the goal of saving this ecosystem and have been recording and reporting on the lake's conditions ever since. In the 2017 Mono Basin Stream Restoration Agreement (MBSRA) was approved, but has not yet implemented by DWP. The green line indicates the lake level that the CA State Water Board mandated must be maintained for the lake to be stable (6,392 feet). The yellow line indicates the level at which land bridges emerge that allow coyotes to prey on nesting birds (6,377 feet above sea level). The orange line indicates the level at which salinity is high enough to reduce brine shrimp populations that many native birds rely on (120 g/L at 6,360 feet above sea level). Point labels indicate significant historical events affecting Mono Lake.