---
title: "Linkages Between Stream Chemistries"
description: |
  An agglomerative heirarchical clustering (via complete linkage) of stream chemistries at the Santa Barbara Coastal LTER.
author:
  - name: Mari Herbst de Cortina
    url: {}
date: 2022-03-01
output:
  distill::distill_article:
    self_contained: false
    code_folding: Click to show code
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# attach packages
library(tidyverse)
library(here)
library(ggdendro)
library(naniar)
library(janitor)
```

### Introduction

In this report I will explore multivariate linkages between water chemistry of streams in the Santa Barbara Coastal LTER. I will use complete linkage analysis to identify clusters among the streams based on the average values of 9 variables: ammonium (uM/L), nitrate (uM/L), phosphate (uM/L), total dissolved nitrogen (um/L), total dissolved phosphorus (uM/L), total particulate carbon (uM/L), total particulate nitrogen (uM/L), total particulate phosphorus (uM/L), total suspended solids (mg/L), and specific conductivity (measured at 25C in sS/cm).

### Complete Agglomerative Heirarchical Clustering

```{r}
# read in data using here to create a system agnostic file path
stream <- read_csv(here("data", "sbc_lter_registered_stream_chemistry.csv")) %>%
  clean_names() %>% # clean up column names
  replace_with_na_all(condition = ~.x == -999) # change -999 values to NA
```

```{r}
# create a subset with summarized values per site
stream_site <- stream %>%
  group_by(site_code) %>%
  summarize(mean_nh4_um = mean(nh4_u_m, na.rm = TRUE), # summarize by the mean value across observations at each site
            mean_no3_um = mean(no3_u_m, na.rm = TRUE),
            mean_po4_um = mean(po4_u_m, na.rm = TRUE),
            mean_tdn_um = mean(tdn_u_m, na.rm = TRUE),
            mean_tdp_um = mean(tdp_u_m, na.rm = TRUE),
            mean_tpc_um = mean(tpc_u_m, na.rm = TRUE),
            mean_tpn_um = mean(tpn_u_m, na.rm = TRUE),
            tss_mgper_liter = mean(tss_mgper_liter, na.rm = TRUE),
            spec_cond_u_spercm = mean(tss_mgper_liter, na.rm = TRUE)) %>%
  drop_na() # drop sites that have NA values
```

```{r}
site_scaled <- stream_site %>%
  select(-site_code) %>% # remove site code column because we don't want to include those values in the analysis
  scale()

rownames(site_scaled) <- stream_site$site_code # add stream site codes back to scaled data as row names
```

```{r}
# find euclidean distance using scaled data
site_dist <- dist(site_scaled,
                  method = "euclidean")

# hierarchical clustering (complete linkage)
site_hc_comp <- hclust(site_dist,
                       method = "complete")
```

#### Dendrogram

```{r}
# save the hierarchical clustering as a dendrogram
site_dend <- as.dendrogram(site_hc_comp)

# visualize
ggdendrogram(site_dend,
             rotate = TRUE) +
  theme_minimal() +
  labs(x = "site ID",
       y = "distance",
       title = "Complete Linkage Dendrogram for Stream Site Chemistry at SBC LTER")
```

**Figure 1**: Complete linkage analysis of clusters between chemistry at streams in the Santa Barbara Coastal LTER. Cluster distance indicates the smallest maximum Euclidean distance between points in the cluster.

### Conclusions

According to the complete linkage dendrogram:

- Stream ON02 is the most unlike the other streams

- MC00 and AB00 are the most similar in their stream chemistry as demonstrated by the short distance between their linkage

### Data Citation

Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 